%%%%%
%%
%% Sample document ``thesis.tex''
%%
%% Version: v0.2
%% Authors: Jean Martina, Rok Strnisa, Matej Urbas
%% Date: 30/07/2008
%%
%% Copyright (c) 2008-2011, Rok Strniša, Jean Martina, Matej Urbas
%% License: Simplified BSD License
%% License file: ./License
%% Original License URL: http://www.freebsd.org/copyright/freebsd-license.html
%%%%%

% Available documentclass options:
%
%   <all `report` document class options, e.g.: `a5paper`>
%   withindex   - enables the index. New index entries can be added through `\index{my entry}`
%   glossary    - enables the glossary.
%   techreport  - typesets the thesis in the technical report format.
%   firstyr     - formats the document as a first-year report.
%   times       - uses the `Times` font.
%   backrefs    - add back references in the Bibliography section
%
% For more info see `README.md`
\documentclass[withindex,glossary,times]{tcd-dissertation}

% Citations using numbers
\usepackage[natbibapa]{apacite}
\bibliographystyle{apacite}
\usepackage{fontspec}
\usepackage{epigraph}
\usepackage{xeCJK}
%\setCJKmainfont{SimSun}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis meta-information
%%

%% The title of the thesis:
\title{English-Irish Machine Translation}

%% The full name of the author (e.g.: James Smith):
\author{Jim O'Regan}

\degreetitle{M.Phil in Speech and Language Processing}
\degreeyear{2017}

\supervisor{Dr. Elaine U\'i Dhonnchadha}

%% Submission date [optional]:
% \submissiondate{November, 2042}

%% You can redefine the submission notice [optional]:
% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}

%% Declaration date:
\date{September, 2017}

%% PDF meta-info:
\subjectline{Irish-English MT}
\keywords{MT, Irish, English}

\foreword{%
This dissertation was supported by funding from The Department of Arts, Heritage, 
Regional, Rural and Gaeltacht Affairs (now The Department of Culture, Heritage and the 
Gaeltacht).

Chapter 2 incorporates text from my contribution to the article ``\guillemotleft All 
Your Data Are Belong to us\guillemotright . European Perspectives on Privacy Issues in 
`Free' Online Machine Translation Services''~\citep{kamocki2015all}, which was carried
out during the period of registration and integrated into the body of the dissertation,
according to the guidelines issued by the Office of the Dean of Graduate Studies,
February 2017, and as discussed with my supervisor.

As this work was created using public funding, and in the interest of openness, I
release this work under the terms of the Creative Commons Attribution 4.0 International 
(CC BY 4.0) licence. As the copyright holder, this notice supercedes any conflicting
notice that may subsequently be put in place by the library of Trinity College, Dublin.
Where such a conflict might occur, the reader is authorised to choose the least 
restrictive of the two.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract:
%%
\abstract{%
This dissertation documents the development of a Rule-Based Machine Translation
system for English to Irish, describing the conversion of a pair of dictionaries
for use as a bilingual lexicon, and the conversion of an existing morphological
analyser for analysis and generation.

It describes a tool created for creating translation rules, based on the alignment
templates used in Statistical Machine Translation. The tool is only partially
complete, as it was intended to also allow the specification of rules with
partial recursion, to reduce the amount of rules that required to be written.
Due to the demands of the English-Irish language pair, rules of a single element
are in the minority; poor foresight on the authors part, along with the complexity
of transforming multiple rule pieces that must remain together, have resulted in
the partial completion of this tool. Consequently, the rules for the English-Irish
pair which were to be described here are not discussed, pending manual expansion
from the form they were written in, which expected this recursive expansion. 

An intended comparison with Statistical Machine Translation and Neural Machine 
Translation was additionally hampered by the poor quality of much of the
available parallel text. This has resulted in the creation of a proof of concept
tool for corpus cleaning, described within.
}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Glossary [optional]:
%%
\newglossaryentry{MT}{
    name=MT,
    description={Machine Translation}
}
\newglossaryentry{EBMT}{
    name=EBMT,
    description={Example-Based Machine Translation}
}
\newglossaryentry{NMT}{
    name=NMT,
    description={Neural Machine Translation}
}
\newglossaryentry{RBMT}{
    name=RBMT,
    description={Rule-Based Machine Translation}
}
\newglossaryentry{SMT}{
    name=SMT,
    description={Statistical Machine Translation}
}
\newglossaryentry{HMM}{
    name=HMM,
    description={Hidden Markov Model}
}
\newglossaryentry{ANN}{
    name=ANN,
    description={Artificial Neural Network}
}
\newglossaryentry{POS}{
    name=POS,
    description={Part of Speech}
}
\newglossaryentry{FST}{
    name=FST,
    description={Finite State Transducer}
}
\newglossaryentry{WFST}{
    name=WFST,
    description={Weighted Finite State Transducer}
}
\newglossaryentry{GPU}{
    name=GPU,
    description={Graphics processing unit}
}
\newglossaryentry{TM}{
    name=TM,
    description={Translation Memory}
}
\newglossaryentry{SRX}{
    name=SRX,
    description={Segmentation Rule eXchange}
}
\newglossaryentry{OCR}{
    name=OCR,
    description={Optical Character Recognition}
}
\newglossaryentry{ASR}{
    name=ASR,
    description={Automatic Speech Recognition}
}
\newglossaryentry{RNN}{
    name=RNN,
    description={Recurrent Neural Network}
}
\newglossaryentry{LSTM}{
    name=LSTM,
    description={Long Short-Term Memory}
}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Contents:
%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgements:
%%
\acknowledgements{%
To my supervisor, Elaine U\'i Dhonnchadha: I know the risk to your reputation
that you took when you accepted me into this programme, and internship, and
that the results documented here cannot vindicate that decision is a bitter,
crushing disappointment. I hope to rectify this as soon as possible. 

To the other members of the 
academic panel: Breffni O'Rourke, Carl Vogel, Christer Gobl, and Ailbhe 
N\'i Chasaide, I am grateful for your roles in making this possible,
and to Kevin Scannell, Piotr Ba\'nski, and Mikel L. Forcada, I am grateful
for the letters of recommendation, and the untiring work each of you do
for the advancement of open source in language technology and resources.

I gratefully acknowledge the support I've received from the Department of 
Culture, Heritage and the Gaeltacht, and thank the Department for it;
without which this work would not have been possible. In particular, 
I'd like to thank Aodh\'an Mac Cormaic: I could not imagine a better
representative for the Department.

To my parents, Joe and Julie: thank you for making me what I am, and for your
support, even at the most trying of times. I couldn't have asked for better
parents.

To my son Mark: you have been my reason to keep going, even when times were
hardest.

To my siblings: Joe, Paul, Angela, Anne Marie, and Catherine: thanks for 
being you. To Joe in particular, thank you for the use of your computer,
without which I could not have completed much of the work I've written about
here. To Anne Marie, as well as my friends James Doyle and Peiyao Li, thank
you for listening to my whining when things were yoyoing out of control.

A debt of gratitude is owed to Joanna Kozicka, for the continuous reminders
that there is a life on the other side of this process. A serious thank
you for sharing your stories -- listening to them is the only thing that
kept me from going completely insane.

The Phonetics and Speech Lab in Trinity has been like a second home for the
past two years, and its other occupants a second family. To M\'iche\'al \'O 
Meachair, Emily Barnes, Fionnuala Nic Ph\'aid\'in, Christoph Wendler, and
Neasa N\'i Chiar\'ain, thank you for your immense help in understanding the
often vague, and frequently underdocumented, aspects of Irish grammar.
Eoghan O'Connor, Andy Murphy, and Maria O'Reilly, thanks for the crosswords.
Jenny O'Reilly, thank you for saving me -- you know how, but it's probably
best not to put it in writing! To Antoin Rodgers (and Eoghan, and James), 
thank you for saving me from so many nights where I could easily have ended
up sleeping rough or roaming the streets. And to all, thanks for the 
doughnuts, the beer, and the many PhLabby memories. (But also, a special
``no thank you'' to Emily for the nickname ``Jimmington'', and my sisters
for adopting it).

To Beata Kozyra: your inspiring me to learn Polish indirectly lead to my
first steps with Natural Language Processing, and watching you painfully
and painstakingly translate a manual, word by word, is what directly lead to my
interest in Machine Translation. Though ours paths may never again cross,
thank you for leading me to the path I'm now following. To Mariusz Florczak
and Tomasz Mikulicz, whom I met on that path, thank you for your help and
your friendship.

Thanks also to my classmates, department colleagues and fellow students over the past two
years. In particular, to Mari Palacios, Vangelis Dimos, Camilla Elan, 
Daniela Modrescu, and Anshu Dash: thanks for your encouragement and support.

A special thank you to Mihael Ar\v{c}an, developer of the IRIS system, and
to Meghan Dowling of Tapad\'oir, for being so keen to share information on
English-Irish Machine Translation.

To the many members of the various open source projects whose software I have
used, and to those with whom I've had the pleasure of collaborating, I thank 
you for your efforts. In particular, the community around Apertium helped
me to get to this point: a special thank you to Fran Tyers, Gema Ram\'irez-S\'anchez,
Sergio Ortiz, Jacob Nordfalk, Kevin Unhammer, Mireia Ginest\'i Rosell, and
Felipe S\'anchez Mart\'inez, as well as to the many students I had the pleasure
of working with as part of Google's Summer of Code and Code-In programmes. I
would here also like to thank my friend and co-author, Pawe\l{} Kamocki, whose
contribution of legal advice does not get quite the amount of recognition it
deserves.

Finally, to ``Oliwka N.'', who (unwittingly) supplied the epigraphs to my chapters:
meeting you was the highest point of the last two years. Muchissimo!

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page, abstract, declaration etc.:
%% -    the title page (is automatically omitted in the technical report mode).
\frontmatter{}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis body:
%%
\include{chapter1}
\include{chapter2}
\include{chapter3}
\include{chapter4}
\include{chapter5}
\include{chapter6}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Bibliography:
%%
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{refs}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix:
%%

\appendix

\include{citinf}
\include{appendixA}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Index:
%%
\printthesisindex

\end{document}
